---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epifish   

<!-- badges: start -->
<!-- badges: end -->

This package provides tools to use Chris Miller's fishplot package (https://github.com/chrisamiller/fishplot) with epidemiological datasets, to generate fishplot epi-curves. 


## Installation

You can install epifish with: 
``` r
#install devtools if you don't have it already for easy installation
install.packages("devtools")
library(devtools)
devtools::install_github("learithe/epifish")
```
  
## Quick demo

Load epifish and required packages
``` r 
library(fishplot); library(dplyr); library(tidyr); library(epifish)
```
```{r, include=FALSE}
library(fishplot); library(dplyr); library(tidyr); library(epifish)
```

Read in the tables of sample data, cluster relationships, and a custom colour scheme
```r
sample_df <- read.csv("epifish/inst/extdata/cluster_counts.csv", stringsAsFactors=FALSE)
parent_df <- read.csv("epifish/inst/extdata/parents.csv", stringsAsFactors=FALSE)
colour_df <- read.csv("epifish/inst/extdata/colours.csv", stringsAsFactors=FALSE)
```

Convert these into a fishplot-ready relative count matrix, fishplot object, and assorted summary data structures:
```{r}
fish_list <- epifish::build_fishplot_tables(sample_df, parent_df, colour_df)
```


Then use the fishplot package to generate the fishplot image:
```{r}
fishplot::fishPlot(fish_list$fish, pad.left=0.1, shape="spline", vlines=fish_list$weeks, vlab=fish_list$weeks)
fishplot::drawLegend(fish_list$fish, nrow=1)

```

Take a look at the underlying matrix and data summaries:
```{r}
fish_list$week_counts

fish_list$week_sums

fish_list$cluster_sums

fish_list$parents

fish_list$raw_table

fish_list$fish_table

fish_list$fish_matrix
```

  
## Input

Example input files/templates can be found in the "inst/extdata" folder in this repository. The basic requirement is a data frame containing one row per sample, with columns `cluster_id` and `week` (any other columns are ignored).

Files can also be provided that describe parent-child relationships for clusters (eg cluster A.1 evolved from cluster A), or a custom colour scheme, but these are optional.

Look at these table structures
```{r}
head(sample_df)

parent_df


colour_df

```


## Output

Output is an R plot. If using RStudio, it is most straightforward to save these as PDF images from the RStudio plot window (Export -> "Save as PDF").



## Notes:

Fishplot citation: Visualizing tumor evolution with the fishplot package for R. Miller CA, McMichael J, Dang HX, Maher CA, Ding L, Ley TJ, Mardis ER, Wilson RK. BMC Genomics. doi:10.1186/s12864-016-3195-z





